An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-19 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More edits to dpp manuscript.

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    \myTheta{C} = \mathbb{E}(\dThetas),
    \label{eq:thetaC}
\end{equation}
where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\dThetas \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units of $\myTheta{C}/\mutationRate$
generations, which I denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be fixed and known according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

\begin{linenomath}
As described by \citet{Oaks2012}, to get the divergence times in units
proportional to the expected number of mutations, I must
scale them by the realized population size for locus $j$ of population-pair $i$ 
\begin{equation}
    \divTimeScaled{i}{j} = \divTimeMap{i} \times \frac{\myTheta{C}}{
        \descendantThetaMean{i} \ploidyScalar{i}{j}},
    \label{eq:divTimeScaled}
\end{equation}
where \descendantThetaMean{i} is the mean of \descendantTheta{1}{} and
\descendantTheta{2}{} for pair $i$.
This gives us the vector of scaled divergence times
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$.
\end{linenomath}

@@

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected four model priors for these analyses (Table~\ref{tabPriors}):
(1) A model to represent the original \msb model, \modelOld, with
priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim U(0,10)$,
$\ancestralTheta{} \sim U(0, 0.05)$,
and
$\descendantThetaMean{} \sim U(0, 0.05)$;
(2) a model with the \msb prior on unordered divergence models, but with
exponential priors on nuisance parameters,
\modelUshaped, with priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$.
(3) a model with uniform prior probability over unordered divergence models,
\modelUniform, with priors
$\divTimeIndexVector \sim \priorUniform$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$;
and
(4) a Dirichlet-process prior model, \modelDPP, with priors
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(2,2)}$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$.
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination,
and unordered divergence models.
For all of my simulations, I used a simulated data structure of eight
population pairs, with a single 1000 base-pair locus sampled from 10
individuals from each population.

>>>>

Unlike the model implemented in \msb, here I place priors on
\divTimeIndexVector and \divTime{}, rather than \divTimeNum and
\divTime{}.
As a result, \divTimeIndexVector determines the number of divergence-time
parameters (\divTimeNum) in the model.
Below, I first describe the prior used for \divTime{} and the timescale it
imposes on the model before discussing the priors implemented for
\divTimeIndexVector.

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    % \myTheta{C} = \mathbb{E}(\dThetas),
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}),
    \label{eq:thetaC}
\end{equation}
% where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$ and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units of $\myTheta{C}/\mutationRate$
generations, which I denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be known and fixed according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

\begin{linenomath}
As described by \citet{Oaks2012}, to get the divergence times in units
proportional to the expected number of mutations, they must be
scaled by the realized population size for locus $j$ of population-pair $i$ 
\begin{equation}
    \divTimeScaled{i}{j} = \divTimeMap{i} \times \frac{\myTheta{C}}{
        \descendantThetaMean{i} \ploidyScalar{i}{j}},
    \label{eq:divTimeScaled}
\end{equation}
where \descendantThetaMean{i} is the mean of \descendantTheta{1}{} and
\descendantTheta{2}{} for pair $i$.
This gives us the vector of scaled divergence times
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$.
\end{linenomath}

@@

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected four model priors for these analyses (Table~\ref{tabPriors}):
(1) A model to represent the original \msb model, \modelOld, with
priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim U(0,10)$,
$\ancestralTheta{} \sim U(0, 0.05)$,
and
$\descendantThetaMean{} \sim U(0, 0.05)$;
(2) a model with the \msb prior on unordered divergence models, but with
exponential priors on nuisance parameters,
\modelUshaped, with priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
(3) a model with uniform prior probability over unordered divergence models,
\modelUniform, with priors
$\divTimeIndexVector \sim \priorUniform$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
and
(4) a Dirichlet-process prior model, \modelDPP, with priors
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(2,2)}$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination,
and unordered divergence models.
For all of my simulations, I used a simulated data structure of eight
population pairs, with a single 1000 base-pair locus sampled from 10
individuals from each population.
